---
title: "Explorative Analysis of Forests and Reforestation Around the Globe"
date: "`r Sys.Date()`"
bibliography: refs.bib
author: "Marcel Schulte"
output: html_document
---
<style>
body {
text-align: justify}
</style>

```{r setup - load data, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, fig.align = "center", dpi = 300, rows.print = 5, cols.min.print = 5)
library(tidyverse)
library(plotly)
library(countrycode)
library(ppsr)
library(shiny)
```

The existence of forests is essential for our life on Earth. By covering around 31 percent of the world’s total land area, one can say that forests are the backbone of entire ecosystems providing a retreat for countless species of animals and plants. A significant part of the oxygen we breathe is provided by the trees, while they also absorb about 25 percent of greenhouse gases. Also economically we are dependent on forests as the livelihoods of about 1.6 billion people around the world are directly or indirectly connected to forests. Despite these utilities, forestation across the world has faced several challenges ranging from wildfires, human-driven deforestation, poor management and poor conservation in general [@unece2019].  

With this project I seek to highlight what has been done by the countries to address these challenges. Therefore I want to figure out and visualize the trends around global reforestation in the last 30 years.  

## Data
The data of the Global Forest Resources assessment (FAO) contains data on forest development for the intervals between 1990 - 2020, which I will use to answer questions around reforestation and deforestation. Furthermore the FAO is providing a dataset which contains forest area for a few intervals. [@fao2020fra].

```{r, include=FALSE}
intervals <- tibble(read_csv("data/fao_fra_datasets/fra_forest_development_data.csv"))
fra <- tibble(read_csv("data/fao_fra_datasets/fra_further_topics_data.csv"))

forest_continents <- read_csv("data/fao_forest_datasets/fao_continents_data.csv", col_types = cols())

colnames(forest_continents)[colnames(forest_continents) == "Value"] <- "ForestSize"
colnames(forest_continents)[colnames(forest_continents) == "Area"] <- "Continent"
```

## Global forest trends

Starting with the current status of our global forests, the trend line of the global forest area gives an overview how the forest area of continents developed. One can see a decrease in the Americas and Africa while Asia recorded a slight increase in forest area. The forest area in Europe as well in Oceania (containing Australia) seems to be quite stable.

Note: the Americas refers to North and South America.

```{r forest-continents-scatter-plot}
newData  %>% 
  ggplot(aes(x = Year, y = ForestSize)) +
  geom_point(aes(color = Continent), size = 3, shape = 16, alpha = 0.7) +
  scale_y_continuous(limits = c(100000, 1800000), labels = scales::comma) +
  labs(
    x = "Year",
    y = "Forest area [1000 ha]",
    title = "Global forest area over the last 30 years"
  ) +
  theme_minimal()
```

The following map is visualizing the change in forest area from 1990 - 2020 for each country. It underlines the previous results. While Brazil and South America in general faces the highest decrease in forest area, China and other Asian countries report a higher value in 2020 compared to 1990.

```{r}
tmp <- fra %>% 
  select(iso3, name, year, `1a_forestArea`) %>% 
  filter(year == 1990 | year == 2020) %>% 
  pivot_wider(names_from = year, values_from = `1a_forestArea`) %>% 
  mutate(diff = `2020` - `1990`)

l <- list(color = toRGB("grey"), width = 0.5)

g <- list(
  showframe = FALSE,
  showcoastlines = FALSE
)

limit <- max(abs(tmp$diff)) * c(-1, 1)

div(plot_ly(tmp, type='choropleth', locations=tmp$iso3, z=tmp$diff, text=tmp$name, colors = "PRGn", marker = list(line = l)) %>%
colorbar(title = "Forest Area\n[1000 ha]", limits = limit) %>%
  layout(
    title = 'Change in forest area from 1990 - 2020',
    geo = g
  ), align = "right")
```

Having now an overview about the global forest area in general I want to dig in deeper into the topic of reforestation by analyzing which countries are the main drivers, whether there is a correlation between reforestation and deforestation and by showing the trends over continents.

However first of all there is some preprocessing needed:

* I summed up the average deforestation and reforestation values from the given intervals to get the total values for each interval and for the whole time period.
* Furthermore I had to change some country names to the short name version, to match the country names to the world map data. 
* I also summed up the reforestation and afforestation values to express any human-made expansion of forest area.

```{r preprocessing_reforestation, include=FALSE}
reforestation <- intervals %>% 
  select(name, year, `1d_afforestation`, `1e_reforestation`) %>% 
  mutate(across(everything(), ~ replace_na(.x, 0))) %>% 
  mutate(`1e_reforestation` = case_when(
    year == c("1990-2000", "2000-2010") ~ `1e_reforestation` * 10,
    TRUE ~ `1e_reforestation` * 5
    ),
    `1d_afforestation` = case_when(
    year == c("1990-2000", "2000-2010") ~ `1d_afforestation` * 10,
    TRUE ~ `1d_afforestation` * 5
    )) %>% 
  mutate(sum = `1d_afforestation` + `1e_reforestation`) 

total_reforestation <- reforestation %>%  
  group_by(name) %>% 
  summarise(totalref = sum(sum, na.rm = TRUE)) %>% 
  ungroup()  %>%
  arrange(desc(totalref))


deforestation <- intervals %>% 
  select(name, year, `1d_deforestation`) %>% 
  mutate(`1d_deforestation` = case_when(
    year == c("1990-2000", "2000-2010") ~ `1d_deforestation` * 10,
    TRUE ~ `1d_deforestation` * 5
    )) %>% 
  group_by(name) %>% 
  summarise(totaldef = sum(`1d_deforestation`, na.rm = TRUE)) %>% 
  ungroup() %>% 
  arrange(desc(totaldef))
```


## Main drivers of reforestation

```{r, include=FALSE}
refplot <- total_reforestation %>%
  head(20)

rest_ref <- total_reforestation %>% 
  slice(21:n()) %>% 
  summarise(name = "Other Countries",
    totalref = sum(totalref))

ref_data <- bind_rows(refplot, rest_ref) %>% mutate(name = factor(name))
```
Following plot shows the Top 20 countries with the highest reforestation between 1990 and 2020.

```{r, echo=FALSE}
 ggplot(ref_data, aes(x = totalref, y = fct_rev(fct_inorder(name)))) +
  geom_col(fill = "Forestgreen") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    x = "Increase in forest area [1000 ha]",
    y = NULL,
    title = "Top 20 countries by total reforestation between 1990 and 2020"
  ) 
```

As these results show mainly countries with a huge surface, I want to put the increase of reforestation from 1990-2020 in relation to the forest area in 1990. 

```{r, include = FALSE}
area <- fra %>%
    group_by(name) %>%
    filter(year == 1990) %>%
    select(name, `1a_forestArea`)
```


```{r, include = FALSE}
increase <- total_reforestation %>%
  inner_join(area, by = "name") %>%
  mutate(reforestation_increase = (totalref/`1a_forestArea`) * 100) %>%
  arrange(desc(reforestation_increase))
```

```{r, echo=FALSE}
increase %>%
    arrange(desc(reforestation_increase)) %>%
    head(20)%>%
  ggplot(mapping = aes(x = reorder(name, reforestation_increase), y = reforestation_increase)) +
  geom_col(fill = "Forestgreen") +
  labs(
    y = "Increase in %",
    x = NULL,
    title = "Countries and their increase in reforestation from 1990-2020\nin relation to their total forest area in 1990"
  ) +
  coord_flip()
```

This leads to quite surprising results, e.g. that Algeria is one of the countries with the highest reforestation increase given the total forest area in 1990. Nevertheless this doesn't come out of nowhere, along with other North African countries, Algeria is pursuing several reforestation projects such as the great green wall or "barrage vert" [@tagesschau]. This results in the fact, that Algeria is one of the countries which has a higher forest cover in 2020 than in 1990.

```{r, include = FALSE, warning = FALSE}
increase$code <- countrycode(increase$name, origin = "country.name", destination = "iso3c")
```

By hovering over the following map, a tooltip with the reforestation increase for each country is shown. The greener the country, the higher is the increase.

```{r, echo=FALSE, warning = FALSE}

l <- list(color = toRGB("grey"), width = 0.5)

g <- list(
  showframe = FALSE,
  showcoastlines = FALSE
)

increase$reforestation_increase <- round(increase$reforestation_increase)

  div(plot_ly(increase, type='choropleth', locations=increase$code, z=increase$reforestation_increase, text=increase$name, colors = "Greens", marker = list(line = l)) %>% 
colorbar(ticksuffix = '%') %>%
  layout(
    title = 'Increase in reforestation between 1990 and 2020 in %', 
    geo = g
  ), align = "right")
```


## Relation between reforestation and deforestation

After we have now an overview on reforestation, I want to answer the question whether governments try to "make up" for their deforestation in the last 30 years.

```{r, include = FALSE}
corref <- total_reforestation %>%
 inner_join(deforestation, by = "name") %>%
 select(name,totalref, totaldef)
```

First of all I want to show the relation of total reforestation and deforestation in the last 30 years, to get a first impression.

```{r, echo=FALSE, warning = FALSE, message = FALSE}
ggplot(data = corref, mapping = aes(x = totalref, y = totaldef)) +
  geom_point(alpha=0.5) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  ggrepel::geom_text_repel(
    data = filter(corref), 
    aes(label = name), 
    color = "black", size = 7/.pt)+
  labs(x = "Total reforestation [1000 ha]", y = "Total deforestation [1000 ha]")
```

There are many outliers with either very high deforestation or reforestation figures. However, such outliers are not surprising when taking a look into recent news. For example Brazil, a country with one of the biggest rainforest areas, has been making negative headlines for years with it's environment politics [@zeitamazonas].

I will downsize the scale of both axis to have a closer look on the data without the outliers.

```{r, echo=FALSE, warning = FALSE, message = FALSE}
ggplot(data = corref, mapping = aes(x = totalref, y = totaldef)) +
  geom_point(alpha=0.5) +
  scale_x_continuous(limits = c(0, 5000), labels = scales::comma) +
  scale_y_continuous(limits = c(0, 10000), labels = scales::comma) +
  ggrepel::geom_text_repel(
    data = filter(corref), 
    aes(label = name), 
    color = "black", size = 7/.pt)+
  labs(x = "Total reforestation [1000 ha]", y = "Total deforestation [1000 ha]")
```

The figure shows already a high and not linear distribution of the data. With the Shapiro-Wilk test I want to show the normality of the data.

```{r, include = FALSE}
corref <- corref %>%
select(totalref, totaldef)
```

```{r, echo=FALSE}
shapiro.test(corref$totalref)
shapiro.test(corref$totaldef)
```

The values are below 0.05 for both, reforestation and deforestation, the data significantly deviate from a normal distribution. A result which is already highlighted by the graph.

As the data is therefore not linear, I choose the Spearman method to calculate the correlation.

```{r, echo=FALSE, warning = FALSE}
c <- cor(x = corref$totalref, y = corref$totaldef, use = "pairwise.complete.obs", method = "spearman")
paste("Spearman = ", c)
```

With a value of 0.54 it shows a strong positive correlation, which means, that deforestation has actually an impact on reforestation and a relationship exists. 

Additionally, I calculated the power predictive score to highlight the impact of deforestation on reforestation. 
 
```{r}
pps <- score_df(corref %>% select(totalref, totaldef))

pps %>% filter(result_type == "predictive power score")
```

However this value is with 0.14 not as high as expected after the correlation result.

## Reforestation trends by continent

In the last step I am going to take a look on the trend line of the continents and their given total reforestation in the given time intervals of 10 years between 1990 and 2020. Again each point represents the total amount of reforestation for the given interval.

```{r, include = FALSE}
world <- map_data("world") %>% 
  group_by(region) %>% 
  filter(row_number()==1)
colnames(world)[5]<-"name"
```

```{r, echo=FALSE, message = FALSE}
trends<- inner_join(world,reforestation,by="name") %>% 
  select(name,year,sum)
trends$continent <- countrycode(sourcevar = trends$name,
                            origin = "country.name",
                            destination = "continent")
trends$year[trends$year=="2010-2015"]<-"2010-2020"
trends$year[trends$year=="2015-2020"]<-"2010-2020"

trends$year[trends$year=="1990-2000"]<-0
trends$year[trends$year=="2000-2010"]<-1
trends$year[trends$year=="2010-2020"]<-2

trends_ref<-trends %>% 
  group_by(year,continent) %>% summarize(t_reforestation =sum(sum,na.rm = TRUE)) %>% ungroup()

ggplot(trends_ref, aes(x = as.integer(year), y = t_reforestation, color = continent)) +
  geom_point(aes(color = continent)) +
  geom_line() +
  scale_x_continuous(breaks = c(0,1,2), labels = c("1990-2000", "2000-2010", "2010-2020")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = 'Reforestation trends on continents between 1990-2020',
    x = 'Year',
    y = 'Reforestation [1000 ha]',
    color = "Continent"
  ) +
  theme_minimal()
```

One can notice, that reforestation in Asia is on a very high level throughout the last 30 years and that Americas are increasing their reforestation efforts with each interval. Also worth mentioning is that Europe does a lot of reforestation in the recent years. It would be very interesting to do another correlation analysis in a few years. 
